Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-06 ind_files/03_field_eqns.tex
Minor phrasing changes, add some references

for every `nice' subset $A \subset \mathcal{T}$. Physicists often write $\delta_{x_i} = \delta(y-x_i)$ as a `function' of a dummy variable $y$ (which will be integrated over). I will stick to the notation $\delta_{x_i}$ because it emphasizes that the object is meant to represent an individual with a trait value of $x_i$ (the dummy variable $y$ can be confusing in this regard). Note that by choosing $f(x) \equiv 1$, we get an `indicator' that is $1$ if the individual is within the set $A$ and $0$ otherwise. Thus, if the population at any time $t$ consists of $N(t)$ individuals with trait values $\{x_1,x_2,\ldots,x_{N(t)}\}$, then it can be completely characterized (Figure \ref{fig_infD_pop_description}) by the `distribution'

@@

    \mathcal{M} = \left\{\sum\limits_{i=1}^{n}\delta_{x_i} \ | \ n \in \mathbb{N}, x_i \in \mathcal{T}\right\}

@@

%     \mathcal{F} = \left\{f:\mathcal{M} \to \mathbb{R} \ | \ f \ \textrm{bounded and measurable}\right\},

@@

\begin{example}
Consider the birth and death functionals:
\begin{equation}
\label{Rogers_logistic_BD}
\begin{aligned}
b(x|\nu) &= r\int\limits_{\mathcal{T}}m(x,y)\nu(y)dy; \ m(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{m}^{2}}\right)\\
    d(x|\nu) &= \frac{\nu(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\nu(y)dy; \ \alpha(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{\alpha}^{2}}\right)
\end{aligned}
\end{equation}
This choice corresponds to an asexual population having a constant (per-capita) birth rate $r$. Birth is sometimes with mutation, and the extent of the mutations is controlled by a Gaussian kernel $m(x,y)$. The death rate is density-dependent, mediated by a Gaussian competition kernel $\alpha(x,y)$, and also contains a phenotype-dependent carrying capacity controlled by $n(x)$, scaled by a constant $K$. The biological interpretation of the death rate is through ecological specialization for limiting resources - Individuals have different intrinsic advantages (controlled by $n(x)$), and experience greater competition from conspecifics that are closer to them in phenotype space (controlled by $\alpha(x,y)$).
\end{example}
Let us now define, for each $x \in \mathcal{T}$, two \emph{step operators} $\mathcal{E}_{x}^{\pm}$ that satisfy
\begin{equation*}
    \mathcal{E}_{x}^{\pm}[f(y,\nu)] =  f(y,\nu \pm \delta_x)
\end{equation*}
In other words, the step operators $\mathcal{E}_{x}^{\pm}$ simply describe the effect of adding or removing a single individual with trait value $x$ from the population.
It is known (only for one-dimensional traits) that we can find a density function $P(\nu,t)$ such that the probability that the process takes value $\nu^{(t)}$ at time $t$ is given by $\int_{\mathcal{T}}P(\nu,t)dx$.\\
We can now use the same trick as earlier and obtain a master equation by counting inflow and outflow of states. Any change to a state must be through the addition or subtraction of a single Dirac delta mass. For any state $\nu \in \mathcal{M}$, the transition rate from $\nu - \delta_{x}$ to $\nu$ is simply $\mathcal{E}^{-}_{x}b(x|\nu)$, and similarly, the transition rate from $\nu+\delta_{x}$ to $\nu$ is $\mathcal{E}^{+}_{x}d(x|\nu)$. The transition rate out of $\nu$ to a state $\nu+\delta_x$ is just $b(x|\nu)$, and transition out to a state $\nu - \delta_x$ is just $d(x|\nu)$. Thus, integrating over all possible $x$ to obtain the total inflow and outflow rate for a state $\nu$, we see that $P(\nu,t)$ must satisfy:
\begin{equation}
\label{unnormalized_M_equation}
    \frac{\partial P}{\partial t}(\nu,t) = \int\limits_{\mathcal{T}}\left[(\mathcal{E}^{-}_{x}-1)b(x|\nu)P(\nu,t) + (\mathcal{E}^{+}_{x}-1)d(x|\nu)P(\nu,t)\right]dx
\end{equation}
This is the `Master equation' of our process.\\
\\
\section{The functional system-size expansion}

To proceed, as before, we assume that there exists a system-size parameter $K > 0$ to obtain a new process $\{\phi^{(t)}\}_{t \geq 0}$ such that for any set $A \subset \mathcal{T}$, $\int_A\phi^{(t)}dx$ gives the `density' of individuals that have trait values that lie within the set $A$. Note that we expect this stochastic process to evolve continuously if $K$ is large since the contribution of each individual is negligible. Specifically, we assume that there exists a $K>0$ such that we can make the substitutions:
\begin{align*}
    \phi^{(t)} = \frac{1}{K}\nu^{(t)} &= \frac{1}{K}\sum\limits_{i=1}^{N(t)}\delta_{x_i}\\
    b_K(x|\phi^{(t)}) &= \frac{1}{K}b(x|\nu^{(t)})\\
    d_K(x|\phi^{(t)}) &= \frac{1}{K}d(x|\nu^{(t)})
\end{align*}
$\{\phi^{(t)}\}_{t\geq0}$ takes values in 
\begin{equation*}
    \mathcal{M}_K = \left\{\frac{1}{K}\sum\limits_{i=1}^{n}\delta_{x_i} \ | \ n \in \mathbb{N}, x_i \in \mathcal{T}\right\}
\end{equation*}
In terms of these new variables, we obtain the master equation:
\begin{equation}
\label{M_equation}
    \frac{\partial P}{\partial t}(\phi,t) = K\int\limits_{\mathcal{T}}\left[(\Delta^{-}_{x}-1)b_K(x|\phi)P(\phi,t) +(\Delta^{+}_{x}-1)d_K(x|\phi)P(\phi,t)\right]dx
\end{equation}
where we have introduced new step operators $\Delta_{x}^{\pm}$ that satisfy:
\begin{equation*}
    \Delta_{x}^{\pm}[F(y,\phi)] =  F\left(y,\phi \pm \frac{1}{K}\delta_x\right)
\end{equation*}
\\
\\
We can now conduct a system-size expansion as before by using a functional `Taylor expansion' of the step operators. Recall that the functional version of the Taylor expansion of a functional $F[\rho]$ about a function $\rho_0$ defined on a domain $\Omega \subseteq \mathbb{R}$ is given by:
\begin{equation*}
    F[\rho_0 + \rho] = F[\rho_0] + \int\limits_{\Omega}\rho(x)\frac{\delta F}{\delta \rho_0(x)}dx + \frac{1}{2!}\int\limits_{\Omega}\int\limits_{\Omega}\rho(x)\rho(y)\frac{\delta^2 F}{\delta \rho_0(x)\delta \rho_0(y)}dxdy + \cdots
\end{equation*}
Since $\Delta^{\pm}_{x}[F[\phi]] = F[\phi \pm \delta_x/K]$, we can Taylor expand the RHS to see that our step operators obey
\begin{align}
    \Delta^{\pm}_{x}[F[\phi]] &= F[\phi] \pm \frac{1}{K}\int\limits_{\mathcal{T}}\frac{\delta F}{\delta \phi(y)}\delta_xdy + \frac{1}{2K^2}\int\limits_{\mathcal{T}}\int\limits_{\mathcal{T}}\frac{\delta^2 F}{\delta \phi(y)\delta \phi(z)}\delta_xdy\delta_xdz+\mathcal{O}(K^{-3})\nonumber\\
    &= F[\phi] \pm \frac{1}{K}\frac{\delta F}{\delta \phi(x)} + \frac{1}{2K^2}\frac{\delta^2 F}{\delta \phi(x)^2}+\mathcal{O}(K^{-3})
    \label{KM_ansatz}
\end{align}
Neglecting terms of $\mathcal{O}(K^{-3})$, we can now substitute \eqref{KM_ansatz} into \eqref{M_equation} to obtain:
\begin{equation*}
\begin{split}
\frac{\partial P}{\partial t}(\phi,t) = K\int\limits_{\mathcal{T}}\left[
    \left(-\frac{1}{K}\frac{\delta}{\delta\phi(x)} + \frac{1}{2K^2}\frac{\delta^2}{\delta\phi(x)^2}\right)\{b_K(x|\phi)P(\phi,t)\}\right]dx\\
    +K\int\limits_{\mathcal{T}}\left[\left(\frac{1}{K}\frac{\delta}{\delta\phi(x)} + \frac{1}{2K^2}\frac{\delta^2}{\delta\phi^2(x)}\right)\{d_K(x|\phi)P(\phi,t)\}\right]dx
\end{split}
\end{equation*}
Rearranging these terms, we obtain a `functional Fokker-Planck equation':
\begin{equation}
\label{functional_FPE}
\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial P}{\partial t}(\phi,t) = \int\limits_{\mathcal{T}}\left[-
    \frac{\delta}{\delta\phi(x)}\{\mathcal{A}^{-}(x|\phi)P(\phi,t)\} + \frac{1}{2K}\frac{\delta^2}{\delta\phi(x)^2}\{\mathcal{A}^{+}(x|\phi)P(\phi,t)\}\right]dx}
\end{equation}
where
\begin{align*}
   \mathcal{A}^{\pm}(x|\phi) &= b_K(x|\phi)\pm d_K(x|\phi) = \frac{1}{K}\left(b(x|\nu)\pm d(x|\nu)\right)
\end{align*}

This constitutes the `mesoscopic' description of our process. We can once again appeal to the link between Fokker-Planck equations and Langevins \citep{lafuerza_role_2016} to say that \eqref{functional_FPE} corresponds to the Langevin equation:
\begin{equation}
\label{functional_langevin}
    \frac{\partial \phi}{\partial t}(x,t) = \mathcal{A}^{-}(x|\phi) + \frac{1}{\sqrt{K}}\eta(x,t)
\end{equation}
where $\eta(x,t)$  is the `Gaussian spacetime white noise' with zero mean and autocovariance function
\begin{equation*}
    \mathbb{E}[\eta(x,t)\eta(x',t')] = \sqrt{\mathcal{A}^{+}(x|\phi)\mathcal{A}^{+}(x'|\phi)}\delta(x-x')\delta(t-t')
\end{equation*}
If one wishes to be mathematically careful, the connection between \eqref{functional_FPE} and \eqref{functional_langevin} becomes somewhat tenuous. In particular, while the equivalence between Fokker-Planck equations and SDEs (Langevin equations) for finite-dimensional stochastic processes is part of the standard mathematical canon, the corresponding equivalence is much less well understood for the infinite-dimensional function-valued processes that we are dealing with, and the interpretation of any formal `Langevin equation' (corresponding now to a stochastic partial differential equation, or SPDE) that we write down is unclear for general measure spaces - We may need $\mathcal{M}_K$ to have a considerable amount of structure (Ex: Seperable Hilbert space) for things to work out \citep{da_prato_stochastic_2014,balan_gentle_2018}, and I do not know whether our domain is nice enough. For example, it is not immediately clear what conditions on $\mathcal{A}^{\pm}(x | \phi)$ are needed for \eqref{functional_langevin} to even admit a solution. Nevertheless, we can recover some well-known deterministic equations from equation \eqref{deterministic_traj}, and under certain assumptions, equations like \eqref{nD_eqn_for_frequencies} have been proved to exist.

@@

b_K(x|\psi) &=  \psi(x,t)b^{\textrm{(ind)}}(x|\psi) + \mu b^{\textrm{(mut)}}(x|\psi)\\

@@

As in chapter \ref{chap_BD}, $ b^{\textrm{(mut)}}(x|\psi)$ describes birth due to mutations and $\mu \geq 0$ is a constant mutation rate. The functions $b^{\textrm{(ind)}}(x|\psi)$ and $d^{\textrm{(ind)}}(x|\psi)$ describe the per-capita birth rate and death rate of type $x$ individuals in a population $\psi$. These functions could in principle model several ecological factors. For example,  $b^{\textrm{(ind)}}(x|\psi)$ may incorporate the effects of mate choice in the sexual case or intrinsic duplication rates in the asexual case, and $d^{\textrm{(ind)}}(x|\psi)$ may model death due to intraspecific competition for resources. Note that the definition of the mean value \eqref{nD_mean} of a type level quantity $f(x)$ now becomes

@@

\frac{\partial \psi}{\partial t}(x,t) = w(x|\psi)\psi(x,t) + \mu b^{\textrm{(mut)}}(x|\psi)

@@

\frac{\partial p}{\partial t} &= \frac{1}{N_K(t)}\left[w(x|\psi)\psi(x,t) + \mu b^{\textrm{(mut)}}(x|\psi)\right] - \frac{\psi(x,t)}{N^2_K(t)}\int\limits_{\mathcal{T}}w(y|\psi)\psi(y,t) + \mu b^{\textrm{(mut)}}(y|\psi)dy\\
&= w(x|\psi)p(x,t) + \frac{\mu}{N_{K}(t)} b^{\textrm{(mut)}}(x|\psi) - p(x,t)\left(\int\limits_{\mathcal{T}}w(y|\psi)p(y,t)dy+\frac{\mu}{N_K(t)}\int\limits_{\mathcal{T}} b^{\textrm{(mut)}}(y|\psi)dy\right)

@@

\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial p}{\partial t}(x,t) = \left[w(x|\psi) - \overline{w}(t)\right]p(x,t)+\frac{\mu}{N_K(t)}\left[b^{\textrm{(mut)}}(x|\psi) - p(x,t)\int\limits_{\mathcal{T}} b^{\textrm{(mut)}}(y|\psi)dy\right]}

@@

This is a continuous version of the replicator-mutator equation when each $x$ is viewed as a strategy. It also yields Kimura's continuum-of-alleles model when each $x$ is viewed as an allele, $b^{\textrm{(mut)}}(x|\psi)$ takes the form of a convolution of $\psi(x,t)$ with a mutation kernel, and the trait space is the entire real line, \emph{i.e.} $\mathcal{T} = \mathbb{R}$. To see this, let $b^{\textrm{(mut)}}(y|\psi) = \int\limits_{\mathbb{R}}m(y-z)\psi(z,t)dz$, where $m(x)$ is a mutation kernel, which by definition is normalized such that $\int\limits_{\mathbb{R}}m(x)dx = 1$. Let us further note that we have implicitly been assuming that the total number of individuals (scaled by $K$) remains finite at all times, \emph{i.e.} $N_K(t) = \int\limits_{\mathbb

@@

&+\frac{\mu}{N_K(t)}\int\limits_{\mathcal{T}}f(x,t)\left[b^{\textrm{(mut)}}(x|\psi) - p(x,t)\int\limits_{\mathcal{T}} b^{\textrm{(mut)}}(y|\psi)dy\right]dx\nonumber\\
&= \overline{fw} - \overline{w}\cdot\overline{f} + \frac{\mu}{N_K(t)}\int\limits_{\mathcal{T}}f(x,t)\left[b^{\textrm{(mut)}}(x|\psi) - p(x,t)\int\limits_{\mathcal{T}} b^{\textrm{(mut)}}(y|\psi)dy\right]dx\label{intermediate_for_cts_price}

@@

M_{\overline{f}}(x|\psi) \coloneqq \frac{\mu}{N_K(t)}\left[\int\limits_{\mathcal{T}}f(x,t)b^{\textrm{(mut)}}(x|\psi)dx - \left(\overline{f}\int\limits_{\mathcal{T}}b^{\textrm{(mut)}}(x|\psi)dx\right)\right] 

@@

    \item Small mutational effects with `almost faithful' reproduction, meaning $b^{\textrm{(mut)}}(x|\psi) \to 0$, and the distribution $\psi(x,t)$ tends to stay very `sharp' (i.e strongly peaked about its mean value). 

@@

\section{Stochastic trait frequency dynamics in the infinite-dimensional case}
In chapter \ref{chap_BD}, we also derived SDEs for the trait frequency dynamics of the complete stochastic case using It\^o's formula. Doing the same for quantitative traits in our framework is tricky because it requires us to find a version of It\^o's formula that holds for SPDEs of the form \eqref{functional_langevin}. The formulation of It\^o formulas and/or an infinite-dimensional stochastic calculus for general function (measure) valued stochastic processes is an active area of research in pure mathematics \citep{da_prato_stochastic_2014}, and without further information on the nature of the domain $\mathcal{M}_K$, it is not clear (to me at least) whether an It\^o's formula exists for our case in general. If $\mathcal{A}^{\pm}(x | \phi)$ in \eqref{functional_FPE} are Gaussian functions, then it has been proven \citep{week_white_2021} that the SDEs `work as expected' if we take $m \to \infty$ in \eqref{nD_eqn_for_frequencies} (see \ref{sec_fun_theorems_mean} and \ref{sec_fun_theorems_var} in chapter \ref{chap_unification}). However, carrying out a general derivation is beyond the scope of this thesis.

@@

Furthermore, for any $\zeta \in \mathcal{M}_K$, we have:

>>>>

for every `nice' subset $A \subset \mathcal{T}$. The Dirac mass is often written $\delta_{x_i} = \delta(y-x_i)$ as a `function' of a dummy variable $y$ which will be integrated over, since in this view the Dirac mass is a `function' that can only occur inside an integral. I will stick to the notation $\delta_{x_i}$ because it emphasizes that $\delta_{x_i}$ is supposed to represent an individual with a trait value of $x_i$ (the dummy variable $y$ can be confusing in this regard). Note that by choosing $f(x) \equiv 1$, we get an `indicator' that is $1$ if the individual is within the set $A$ and $0$ otherwise. Thus, if the population at any time $t$ consists of $N(t)$ individuals with trait values $\{x_1,x_2,\ldots,x_{N(t)}\}$, then it can be completely characterized (Figure \ref{fig_infD_pop_description}) by the `distribution'

@@

    \mathcal{M}(\mathcal{T})= \left\{\sum\limits_{i=1}^{n}\delta_{x_i} \ | \ n \in \mathbb{N}, x_i \in \mathcal{T}\right\}

@@

%     \mathcal{F} = \left\{f:\mathcal{M}(\mathcal{T})\to \mathbb{R} \ | \ f \ \textrm{bounded and measurable}\right\},

@@

\begin{example}
Consider the birth and death functionals:
\begin{equation}
\label{Rogers_logistic_BD}
\begin{aligned}
b(x|\nu) &= r\int\limits_{\mathcal{T}}m(x,y)\nu(y)dy; \ m(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{m}^{2}}\right)\\
    d(x|\nu) &= \frac{\nu(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\nu(y)dy; \ \alpha(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{\alpha}^{2}}\right)
\end{aligned}
\end{equation}
This choice corresponds to an asexual population having a constant (per-capita) birth rate $r$. Birth is sometimes with mutation, and the extent of the mutations is controlled by a Gaussian kernel $m(x,y)$. The death rate is density-dependent, mediated by a Gaussian competition kernel $\alpha(x,y)$, and also contains a phenotype-dependent carrying capacity controlled by $n(x)$, scaled by a constant $K$. The biological interpretation of the death rate is through ecological specialization for limiting resources - Individuals have different intrinsic advantages (controlled by $n(x)$), and experience greater competition from conspecifics that are closer to them in phenotype space (controlled by $\alpha(x,y)$).
\end{example}
Let us now define, for each $x \in \mathcal{T}$, two \emph{step operators} $\mathcal{E}_{x}^{\pm}$ that satisfy
\begin{equation*}
    \mathcal{E}_{x}^{\pm}[f(y,\nu)] =  f(y,\nu \pm \delta_x)
\end{equation*}
In other words, the step operators $\mathcal{E}_{x}^{\pm}$ simply describe the effect of adding or removing a single individual with trait value $x$ from the population.
It is known (only for one-dimensional traits) that we can find a density function $P(\nu,t)$ such that the probability that the process takes value $\nu^{(t)}$ at time $t$ is given by $\int_{\mathcal{T}}P(\nu,t)dx$.\\
We can now use the same trick as earlier and obtain a master equation by counting inflow and outflow of states. Any change to a state must be through the addition or subtraction of a single Dirac delta mass. For any state $\nu \in \mathcal{M}(\mathcal{T})$, the transition rate from $\nu - \delta_{x}$ to $\nu$ is simply $\mathcal{E}^{-}_{x}b(x|\nu)$, and similarly, the transition rate from $\nu+\delta_{x}$ to $\nu$ is $\mathcal{E}^{+}_{x}d(x|\nu)$. The transition rate out of $\nu$ to a state $\nu+\delta_x$ is just $b(x|\nu)$, and transition out to a state $\nu - \delta_x$ is just $d(x|\nu)$. Thus, integrating over all possible $x$ to obtain the total inflow and outflow rate for a state $\nu$, we see that $P(\nu,t)$ must satisfy:
\begin{equation}
\label{unnormalized_M_equation}
    \frac{\partial P}{\partial t}(\nu,t) = \int\limits_{\mathcal{T}}\left[(\mathcal{E}^{-}_{x}-1)b(x|\nu)P(\nu,t) + (\mathcal{E}^{+}_{x}-1)d(x|\nu)P(\nu,t)\right]dx
\end{equation}
This is the `Master equation' of our process.\\
\\
\section{The functional system-size expansion}

To proceed, as before, we assume that there exists a system-size parameter $K > 0$ to obtain a new process $\{\phi^{(t)}\}_{t \geq 0}$ such that for any set $A \subset \mathcal{T}$, $\int_A\phi^{(t)}dx$ gives the `density' of individuals that have trait values that lie within the set $A$. Note that we expect this stochastic process to evolve continuously if $K$ is large since the contribution of each individual is negligible. Specifically, we assume that there exists a $K>0$ such that we can make the substitutions:
\begin{align*}
    \phi^{(t)} = \frac{1}{K}\nu^{(t)} &= \frac{1}{K}\sum\limits_{i=1}^{N(t)}\delta_{x_i}\\
    b_K(x|\phi^{(t)}) &= \frac{1}{K}b(x|\nu^{(t)})\\
    d_K(x|\phi^{(t)}) &= \frac{1}{K}d(x|\nu^{(t)})
\end{align*}
$\{\phi^{(t)}\}_{t\geq0}$ takes values in 
\begin{equation*}
    \mathcal{M}_{K}(\mathcal{T}) = \left\{\frac{1}{K}\sum\limits_{i=1}^{n}\delta_{x_i} \ | \ n \in \mathbb{N}, x_i \in \mathcal{T}\right\}
\end{equation*}
In terms of these new variables, we obtain the master equation:
\begin{equation}
\label{M_equation}
    \frac{\partial P}{\partial t}(\phi,t) = K\int\limits_{\mathcal{T}}\left[(\Delta^{-}_{x}-1)b_K(x|\phi)P(\phi,t) +(\Delta^{+}_{x}-1)d_K(x|\phi)P(\phi,t)\right]dx
\end{equation}
where we have introduced new step operators $\Delta_{x}^{\pm}$ that satisfy:
\begin{equation*}
    \Delta_{x}^{\pm}[F(y,\phi)] =  F\left(y,\phi \pm \frac{1}{K}\delta_x\right)
\end{equation*}
\\
\\
We can now conduct a system-size expansion as before by using a functional `Taylor expansion' of the step operators. Recall that the functional version of the Taylor expansion of a functional $F[\rho]$ about a function $\rho_0$ defined on a domain $\Omega \subseteq \mathbb{R}$ is given by:
\begin{equation*}
    F[\rho_0 + \rho] = F[\rho_0] + \int\limits_{\Omega}\rho(x)\frac{\delta F}{\delta \rho_0(x)}dx + \frac{1}{2!}\int\limits_{\Omega}\int\limits_{\Omega}\rho(x)\rho(y)\frac{\delta^2 F}{\delta \rho_0(x)\delta \rho_0(y)}dxdy + \cdots
\end{equation*}
Since $\Delta^{\pm}_{x}[F[\phi]] = F[\phi \pm \delta_x/K]$, we can Taylor expand the RHS to see that our step operators obey
\begin{align}
    \Delta^{\pm}_{x}[F[\phi]] &= F[\phi] \pm \frac{1}{K}\int\limits_{\mathcal{T}}\frac{\delta F}{\delta \phi(y)}\delta_xdy + \frac{1}{2K^2}\int\limits_{\mathcal{T}}\int\limits_{\mathcal{T}}\frac{\delta^2 F}{\delta \phi(y)\delta \phi(z)}\delta_xdy\delta_xdz+\mathcal{O}(K^{-3})\nonumber\\
    &= F[\phi] \pm \frac{1}{K}\frac{\delta F}{\delta \phi(x)} + \frac{1}{2K^2}\frac{\delta^2 F}{\delta \phi(x)^2}+\mathcal{O}(K^{-3})
    \label{KM_ansatz}
\end{align}
Neglecting terms of $\mathcal{O}(K^{-3})$, we can now substitute \eqref{KM_ansatz} into \eqref{M_equation} to obtain:
\begin{equation*}
\begin{split}
\frac{\partial P}{\partial t}(\phi,t) = K\int\limits_{\mathcal{T}}\left[
    \left(-\frac{1}{K}\frac{\delta}{\delta\phi(x)} + \frac{1}{2K^2}\frac{\delta^2}{\delta\phi(x)^2}\right)\{b_K(x|\phi)P(\phi,t)\}\right]dx\\
    +K\int\limits_{\mathcal{T}}\left[\left(\frac{1}{K}\frac{\delta}{\delta\phi(x)} + \frac{1}{2K^2}\frac{\delta^2}{\delta\phi^2(x)}\right)\{d_K(x|\phi)P(\phi,t)\}\right]dx
\end{split}
\end{equation*}
Rearranging these terms, we obtain a `functional Fokker-Planck equation':
\begin{equation}
\label{functional_FPE}
\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial P}{\partial t}(\phi,t) = \int\limits_{\mathcal{T}}\left[-
    \frac{\delta}{\delta\phi(x)}\{\mathcal{A}^{-}(x|\phi)P(\phi,t)\} + \frac{1}{2K}\frac{\delta^2}{\delta\phi(x)^2}\{\mathcal{A}^{+}(x|\phi)P(\phi,t)\}\right]dx}
\end{equation}
where
\begin{align*}
   \mathcal{A}^{\pm}(x|\phi) &= b_K(x|\phi)\pm d_K(x|\phi) = \frac{1}{K}\left(b(x|\nu)\pm d(x|\nu)\right)
\end{align*}

This constitutes the `mesoscopic' description of our process. We can once again appeal to the link between Fokker-Planck equations and Langevins \citep{lafuerza_role_2016} to say that \eqref{functional_FPE} corresponds to the Langevin equation:
\begin{equation}
\label{functional_langevin}
    \frac{\partial \phi}{\partial t}(x,t) = \mathcal{A}^{-}(x|\phi) + \frac{1}{\sqrt{K}}\eta(x,t)
\end{equation}
where $\eta(x,t)$  is the `Gaussian spacetime white noise' with zero mean and autocovariance function
\begin{equation*}
    \mathbb{E}[\eta(x,t)\eta(x',t')] = \sqrt{\mathcal{A}^{+}(x|\phi)\mathcal{A}^{+}(x'|\phi)}\delta(x-x')\delta(t-t')
\end{equation*}
If one wishes to be mathematically careful, the connection between \eqref{functional_FPE} and \eqref{functional_langevin} becomes somewhat tenuous. In particular, while the equivalence between Fokker-Planck equations and SDEs (Langevin equations) for finite-dimensional stochastic processes is part of the standard mathematical canon, the corresponding equivalence is much less well understood for the infinite-dimensional function-valued processes that we are dealing with, and the interpretation of any formal `Langevin equation' (corresponding now to a stochastic partial differential equation, or SPDE) that we write down is unclear for general measure spaces - We may need $\mathcal{M}_{K}(\mathcal{T})$ to have a considerable amount of structure (Ex: Seperable Hilbert space) for things to work out \citep{da_prato_stochastic_2014,balan_gentle_2018}, and I do not know whether our domain is nice enough. For example, it is not immediately clear what conditions on $\mathcal{A}^{\pm}(x | \phi)$ are needed for \eqref{functional_langevin} to even admit a solution. Nevertheless, we can recover some well-known deterministic equations from equation \eqref{deterministic_traj}, and under certain assumptions, equations like \eqref{nD_eqn_for_frequencies} have been proved to exist.

@@

b_K(x|\psi) &=  \psi(x,t)b^{\textrm{(ind)}}(x|\psi) + \mu Q(x|\psi)\\

@@

As in chapter \ref{chap_BD}, $ Q(x|\psi)$ describes birth due to mutations and $\mu \geq 0$ is a constant mutation rate. The functions $b^{\textrm{(ind)}}(x|\psi)$ and $d^{\textrm{(ind)}}(x|\psi)$ describe the per-capita birth rate and death rate of type $x$ individuals in a population $\psi$. These functions could in principle model several ecological factors. For example,  $b^{\textrm{(ind)}}(x|\psi)$ may incorporate the effects of mate choice in the sexual case or intrinsic duplication rates in the asexual case, and $d^{\textrm{(ind)}}(x|\psi)$ may model death due to intraspecific competition for resources. Note that the definition of the mean value \eqref{nD_mean} of a type level quantity $f(x)$ now becomes

@@

\frac{\partial \psi}{\partial t}(x,t) = w(x|\psi)\psi(x,t) + \mu Q(x|\psi)

@@

\frac{\partial p}{\partial t} &= \frac{1}{N_K(t)}\left[w(x|\psi)\psi(x,t) + \mu Q(x|\psi)\right] - \frac{\psi(x,t)}{N^2_K(t)}\int\limits_{\mathcal{T}}w(y|\psi)\psi(y,t) + \mu b^{\textrm{(mut)}}(y|\psi)dy\\
&= w(x|\psi)p(x,t) + \frac{\mu}{N_{K}(t)} Q(x|\psi) - p(x,t)\left(\int\limits_{\mathcal{T}}w(y|\psi)p(y,t)dy+\frac{\mu}{N_K(t)}\int\limits_{\mathcal{T}} b^{\textrm{(mut)}}(y|\psi)dy\right)

@@

\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial p}{\partial t}(x,t) = \left[w(x|\psi) - \overline{w}(t)\right]p(x,t)+\frac{\mu}{N_K(t)}\left[Q(x|\psi) - p(x,t)\int\limits_{\mathcal{T}} b^{\textrm{(mut)}}(y|\psi)dy\right]}

@@

This is a continuous version of the replicator-mutator equation when each $x$ is viewed as a strategy. It also yields Kimura's continuum-of-alleles model when each $x$ is viewed as an allele, $Q(x|\psi)$ takes the form of a convolution of $\psi(x,t)$ with a mutation kernel, and the trait space is the entire real line, \emph{i.e.} $\mathcal{T} = \mathbb{R}$. To see this, let $b^{\textrm{(mut)}}(y|\psi) = \int\limits_{\mathbb{R}}m(y-z)\psi(z,t)dz$, where $m(x)$ is a mutation kernel, which by definition is normalized such that $\int\limits_{\mathbb{R}}m(x)dx = 1$. Let us further note that we have implicitly been assuming that the total number of individuals (scaled by $K$) remains finite at all times, \emph{i.e.} $N_K(t) = \int\limits_{\mathbb

@@

&+\frac{\mu}{N_K(t)}\int\limits_{\mathcal{T}}f(x,t)\left[Q(x|\psi) - p(x,t)\int\limits_{\mathcal{T}} b^{\textrm{(mut)}}(y|\psi)dy\right]dx\nonumber\\
&= \overline{fw} - \overline{w}\cdot\overline{f} + \frac{\mu}{N_K(t)}\int\limits_{\mathcal{T}}f(x,t)\left[Q(x|\psi) - p(x,t)\int\limits_{\mathcal{T}} b^{\textrm{(mut)}}(y|\psi)dy\right]dx\label{intermediate_for_cts_price}

@@

M_{\overline{f}}(x|\psi) \coloneqq \frac{\mu}{N_K(t)}\left[\int\limits_{\mathcal{T}}f(x,t)Q(x|\psi)dx - \left(\overline{f}\int\limits_{\mathcal{T}}Q(x|\psi)dx\right)\right] 

@@

    \item Small mutational effects with `almost faithful' reproduction, meaning $Q(x|\psi) \to 0$, and the distribution $\psi(x,t)$ tends to stay very `sharp' (i.e strongly peaked about its mean value). 

@@

\section{Stochastic trait frequency dynamics in the infinite-dimensional case}
In chapter \ref{chap_BD}, we also derived SDEs for the trait frequency dynamics of the complete stochastic case using It\^o's formula. Doing the same for quantitative traits in our framework is tricky because it requires us to find a version of It\^o's formula that holds for SPDEs of the form \eqref{functional_langevin}. The formulation of It\^o formulas and/or an infinite-dimensional stochastic calculus for general function (measure) valued stochastic processes is an active area of research in pure mathematics \citep{da_prato_stochastic_2014}, and without further information on the nature of the domain $\mathcal{M}_{K}(\mathcal{T})$, it is not clear (to me at least) whether an It\^o's formula exists for our case in general. If $\mathcal{A}^{\pm}(x | \phi)$ in \eqref{functional_FPE} are Gaussian functions, then it has been proven \citep{week_white_2021} that the SDEs `work as expected' if we take $m \to \infty$ in \eqref{nD_eqn_for_frequencies} (see \ref{sec_fun_theorems_mean} and \ref{sec_fun_theorems_var} in chapter \ref{chap_unification}). However, carrying out a general derivation is beyond the scope of this thesis.

@@

Furthermore, for any $\zeta \in \mathcal{M}_{K}(\mathcal{T})$, we have:
